---
title: "Class10"
author: "Patricia Chen A16138722"
format: pdf
editor: visual
---

# 4. Comparative structure analysis of Adenylate Kinase (ADK)

ADK is am important drug target and we would love to know how it works ie. molecular mechanism. There has been a lot of work done on this protein due to it's importance inclusing lots of crustal structures.

We will begin with getting an example ADK sequence from the database, we will then use this to find all ADK structures in the pDB

## Install packages in the R console NOT your Rmd/Quarto file

```{r}

#install.packages("bio3d")
#install.packages("devtools")
#install.packages("BiocManager")

#BiocManager::install("msa")
#devtools::install_bitbucket("Grantlab/bio3d-view")

```

> Q10. Which of the packages above is found only on BioConductor and not CRAN? 
> Answer: MSA

> Q11. Which of the above packages is not found on BioConductor or CRAN?:
> Answer: The bottom package, bio3d-view, is not found on the BioConductor or CRAN.

> Q12. True or False? Functions from the devtools package can be used to install packages from GitHub and BitBucket?
> Answer: True.

## Search and retrieve ADK structures

```{r}
library(bio3d)
aa <- get.seq("1ake_A")

aa
```

We can now run BLAST

```{r}
#b<- blast.pdb(aa)
```

```{r}
#hits<-plot(b)
```

Let's see what is in our 'hits' object
```{r}
#hits$pdb.id

```

hits$pdb.id


> Q13. How many amino acids are in this sequence, i.e. how long is this sequence? 
> Answer: There are 214 amino acids in this sequence.


```{r}
# Blast or hmmer search 
#b <- blast.pdb(aa)

# Plot a summary of search results
#hits <- plot(b)

# List out some 'top hits'
#head(hits$pdb.id)

hits <- NULL
hits$pdb.id <- c('1AKE_A','6S36_A','6RZE_A','3HPR_A','1E4V_A','5EJE_A','1E4Y_A','3X2S_A','6HAP_A','6HAM_A','4K46_A','3GMT_A','4PZL_A')
```

We can now use function get.pdb() and pdbslit() to fetch and parse the identified structures.

```{r}
# Download releated PDB files - Skipping downloads (already exists)
files <- get.pdb(hits$pdb.id, path="pdbs", split=TRUE, gzip=TRUE)

```

# Alighment: Align and superpose structures
Next we will use the pdbaln() function to align and also optionally fit (i.e. superpose) the identified PDB structures.
Now I want to align and supersose theses structures which are all over the place

```{r}

# Align releated PDBs
pdbs <- pdbaln(files, fit = TRUE, exefile="msa")

```

Let's have a look at our 'pdbs' object
```{r}
pdbs
```

Now we have our aligned and superposed structures we can perform all sorts of analysis on them. Let's do PCA..


```{r}
# Vector containing PDB codes for figure axis
#ids <- basename.pdb(pdbs$id)

# Draw schematic alignment
#plot(pdbs, labels=ids)
```

## Annotate collected PDB structures

```{r}
#anno <- pdb.annotate(ids)
#unique(anno$source)
#anno
```

Principal component analysis

```{r}
# Perform PCA
pc.xray <- pca(pdbs)
plot(pc.xray)
```


Results of PCA on ADK X-ray structure. Each dot represents one PDB structure.

We can cluster the structures by RMSD (or any other methods).

```{r}
# Calculate RMSD, measuring the distance b/w clusters
rd <- rmsd(pdbs)

# Structure-based clustering
hc.rd <- hclust(dist(rd))
grps.rd <- cutree(hc.rd, k=3)


#plot(pc.xray, 1:2, col=grps)
plot(pc.xray, 1:2, col="grey50", bg=grps.rd, pch=21, cex=1)
```

# 5. Optional further visualization

We can make a movie - also called a trajectory of the major differences (ie. structural displacement) of ADK

```{r}
# Visualize first principal component
pc1 <- mktrj(pc.xray, pc=1, file="pc_1.pdb")
```




